DataCamp: An Interactive Learning Tool for Programming

### 2018-04-09 manuscript.tex

\usepackage{soul}     % for highlighting
\usepackage{upquote}  % for correct display of single quotes in verbatim
\usepackage{booktabs} % for toprule, etc. in tabular

@@

\title{DataCamp: An Interactive Teaching Tool for Programming}

\author{Tessa Mcdonnel, Mine \c{C}etinkaya-Rundel, Weston Stearns, Shannon Pileggi}

@@

@@

With the increased use of R in classrooms, an effective tool to familiarize students
with statistical software is needed \citep{Baumer2014}. While R software provides users
with extraordinary statistical capabilities, there is certainly a steep learning curve.
To facilitate this transition to R programming, popular interactive teaching technologies
have been created including \texttt{swirl} \citep{Kross}, Try R \citep{TryR}, LearnR, and DataCamp.
Interactive tools allow students to actively work through coding problems and receive
immediate feedback.

%--------------------------------------------------------------------------------
\subsubsection{Swirl}

\texttt{Swirl} is an open source R package where students can learn R programming
directly in the R console \citep{Kross}. This interactive learning tool contains several
pre-existing lessons but educators can also create their own lesson content to fit
specific curriculum \citep{Carchedi2014}.  While \textit{swirl} can be linked to Massive
Open Online Courses (MOOCs) like Coursera, integration of learning management systems (LMS)
is not currently available.

@@

Try R is another interactive R programming course but unlike \texttt{swirl}, Try R
is web-based \citep{TryR}. The RStudio console can be intimidating to new statistics students
so an  advantage of a web-based teaching tool is the clean and easy to use interface.
User-friendliness is an important quality given that students often struggle more
with the software used than the statistical concepts themselves \citep{Hare2017}.
When students only need their web browser to participate, all installation
complications are avoided and valuable class time can be focused on the lesson.
However, Try R is not open source so public content authoring is not available;
educators are limited to the existing lessons available on the Try R website.
Lastly, Try R cannot currently integrate with LMS.

%--------------------------------------------------------------------------------
\subsubsection{LearnR?}

@@

The simplest way to begin building the course is to use the template course files from
the \texttt{Create Course} dialog. This will automatically create a GitHub repository
that is linked to DataCamp.

@@

Course creators can edit their courses through either GitHub or the \texttt{Teach Editor}
accessed through the DataCamp website. The \texttt{Teach Editor} allows content
creators to edit, preview, save changes, and synchronize to the GitHub repository.
Editing completed via GitHub does not allow for automatic content viewing. Any changes
that are made through the DataCamp \texttt{Teach Editor} will update the GitHub
repository and vice versa.

@@

Figure \ref{fig:code1} is an example of ``normal exercise'' syntax and Figure 
\ref{fig:preview} is the corresponding interface students see. Students read the 
lesson information and instructions, answer the question, and submit their code.

@@

\begin{figure}
\caption{Syntax to create a ``normal exercise'' on DataCamp.  In DataCamp's
\texttt{Teach Editor}, the syntax is color-coded.}
\begin{Verbatim}[frame=single]
--- type:NormalExercise lang:r xp:100 skills:1 key:9a4fea2da0

@@

\end{Verbatim}
\label{fig:code1}
\end{figure}

\begin{figure}[h] 
\caption{Preview of the rendered result of the ``normal exercise'' syntax displayed 
in Figure \ref{fig:code1}.}
\includegraphics[width = 1.0\textwidth] {code1preview.png}
\label{fig:preview}
\end{figure}

@@

The SCT automatically assesses if the correct code was submitted. If a student submits 
an incorrect answer, they will receive immediate feedback and guidance towards
the correct answer.  The SCT is executed through DataCamp's \texttt{testwhat} R
package (\url{https://github.com/datacamp/testwhat/wiki}). The \textit{testwhat} package 
contains functions that test different types of problems including multiple choice
exercises, what students typed, and output generated.  The SCT compares the ideal answer 
(from the solution code) to the student's answer. Various arguments can be added to the 
test functions to adjust the testing process and feedback.  Instructors can either use the 
default feedback that DataCamp automatically generates or write their own custom messages. 
For example, Table \ref{tab:SCT} contains the default feedback for a few possible incorrect submissions 
of the \texttt{hist()} function from the example
exercise \textit{Visualizing quantitative data in R} shown in Figures \ref{fig:code1} and
\ref{fig:preview}.

\begin{table}\label{tab:SCT}
\caption{Feedback provided in the R console and in the DataCamp lesson through the 
SCT for the exercise shown in Figures \ref{fig:code1} and \ref{fig:preview}.}
\begin{tabular}{lll}
\toprule
Student submission & R console & SCT feedback \\
\midrule
\texttt{histogram(CPS85\$wage)} & Error: could not find & Have you called hist()? \\
                                & function ``histogram'' & \\
[2ex]
\texttt{hist(wage)} & Error: object `wage' & Follow the format: hist(dataset\$variable) \\
                    &  not found           & with specified dataset and variable.\\
[2ex]                    
\texttt{hist(CP85\$wage)} & Error: object `CP85' & Follow the format: hist(dataset\$variable) \\
                               & not found       & with specified dataset and variable. \\
[2ex]                                                   
\texttt{hist(CPS85\$wage)} & & Great work! \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}\label{tab:exercises}
\caption{Exercise types currently supported in DataCamp}
\begin{tabular}{ll}
\toprule
Type & Description \\
\midrule
\texttt{VideoExercise}              & Displays a video exercise \\
\texttt{NormalExercise}	            & Instructions, exercise, code editor, and console \\
\texttt{MultipleChoiceExercise}     & Multiple choice question and console \\
\texttt{PureMultipleChoiceExercise} & Multiple choice question without a console \\
\texttt{TabExercise}	            & A series of connected sub-exercises \\
\texttt{BulletExercise}	            & A series of connected sub-exercises \\
\bottomrule
\end{tabular}
\end{table}

@@

The \texttt{MultipleChoiceExercise} contains a question and a list of various answers. 
This exercise is accompanied by the R console and R output, and allows instructors to 
pre-program objects or plots that the student needs to access to answer the question.
The \texttt{PureMultipleChoiceExercise} contains a question and a list of various
answers - this exercise type does not include the R console and output. Similar to a
\texttt{NormalExercise}, the lesson, instructions, hint, pre-exercise code and SCTs are 
included in the exercise syntax. Instructors write the questions in the lesson portion 
of the exercise, present options in the \textit{Instructions} section, and can specify a 
\textit{hint} as well. The \texttt{MultipleChoiceExercise} utilizes SCT for assessment,
whereas the \texttt{PureMultipleChoiceExercise} utilizes simpler syntax.

@@

Each DataCamp course is associated with a primary content creator, but multiple individuals 
can contribute simultaneously to the course.  To do so, go to the course repository on GitHub 
and under \texttt{Settings} enter the Github user names of collaborators. One note of caution 
here - to deploy the course to your class, the primary content creator needs to be listed 
as member of the course.

@@

Integration of DataCamp assignments with Moodle was easy.  A link with the assigned
chapter was posted on the course site, and then students were taken directly to the 
first page of the assigned chapter (no additional log in required!). Grades were 
available in the Moodle grade book immediately after students completed the assignment.
DataCamp computes scores as experience points (\texttt{xp}), but Moodle automatically 
converted this a percentage completion as desired.  In addition, students were
able to re-visit the chapter as many times as they wanted.

@@

While there will always be some students (among non-statistics majors) that grumble about 
having to learn R, overall student feedback was positive.

@@

With integration via Moodle, one chapter was assigned at a time.  Although it
was possible to specify a recommended deadline for assignment completion, this
was not enforced via Moodle.  This meant that students could complete the
assignment after the deadline and still receive full credit.  Moreover,
because students are able to re-visit a chapter, the grade recorded in the Moodle
grade center was always the highest grade recorded for the student, regardless of
the completion iteration (first, second attempt?) or date of completion.

@@

@@

-The only disadvantage of the modularity feature is that the sessions are not cumulative; every
exercise begins with a clear work space so an object defined in a previous exercise needs
to be defined again.

-DataCamp is changing constantly

-emphasis on correct code over interpretation of results

-no way to produce reproducible reports based on data camp work

-no notes/summaries provided - easy to forget

-in the classroom

-glitches in teach editor

@@

\bibliographystyle{plainnat}
\bibliography{library}

>>>>

\usepackage{soul}     % for highlighting
\usepackage{upquote}  % for correct display of single quotes in verbatim
\usepackage{booktabs} % for toprule, etc. in tabular
\usepackage{pifont}   % used to create check/x marks

\newcommand{\cmark}{\ding{51}} % check mark
\newcommand{\xmark}{\ding{55}} % x mark

@@

\title{DataCamp: An Interactive Learning Tool for Programming}

@@

\hl{Discuss benefits of `learn by doing' or repetition?  Ie, this incorporates
more student practice with typing code.}

@@

With the increased use of R in classrooms, an effective tool to familiarize students
with statistical software is needed \citep{Baumer2014}. While R software provides users
with extraordinary statistical capabilities, there is certainly a steep learning curve.
To facilitate this transition to R programming, popular interactive teaching technologies
have been created including Try R, \texttt{swirl}, \texttt{learnr}, and DataCamp.
Interactive tools allow students to actively work through coding problems and receive
immediate feedback.  See Table \ref{tab:compare} for a comparison of existing
technologies.

\hl{Best placement for this table?}

\begin{table}
\begin{tabular}{lcccc}
\toprule
Feature                      & Try R  & \texttt{swirl}    & \texttt{learnr} & DataCamp \\
\midrule
%Requires R installation     & \cmark & \cmark            &  $\sim$         & \xmark   \\
Web-based                    & \cmark & \xmark            &  $\sim$         & \cmark   \\
Tracks student progress      & \xmark & $\sim$            &  \xmark         & \cmark   \\
Integrates with LMS          & \xmark & \xmark            &  \xmark         & \cmark   \\
Variety of lessons available & \xmark & \cmark            &  \xmark         & \cmark   \\
Educators can create content & \xmark & \cmark            &  \cmark         & \cmark   \\
%Meta-data available         & \xmark &   ?               &                 & \cmark   \\
\bottomrule
\end{tabular}
\caption{Comparison of interactive technologies to learn programming in R.
For web-based applications, R installation is not required.}
\label{tab:compare}
\end{table}

@@

Try R is web-based interactive R programming course that consists only of
codding exercises \citep{TryR}. The RStudio console
can be intimidating to new statistics students
so an  advantage of a web-based teaching tool is the clean and easy to use interface.
User-friendliness is an important quality given that students often struggle more
with the software used than the statistical concepts themselves \citep{Hare2017}.
When students only need their web browser to participate, all installation
complications are avoided and valuable class time can be focused on the lesson.
However, Try R is not open source so public content authoring is not available;
educators are limited to the existing lessons available on the Try R website.
Lastly, Try R does not currently integrate with learning management systems (LMS).

%--------------------------------------------------------------------------------
\subsubsection{swirl}

\texttt{swirl} is an open source R package where students can learn R programming
directly in the R console (so it requires a local R installation) \citep{Kross}.
This interactive learning tool incorporates both code exercises and multiple choice
questions. \texttt{swirl} contains several pre-existing lessons, and
educators can also create their own lesson content to fit
specific curriculum \citep{Carchedi2014}.  While \texttt{swirl} can be linked to Massive
Open Online Courses (MOOCs) like Coursera, integration of LMS
is not currently available.

%--------------------------------------------------------------------------------
\subsubsection{learnr}

\texttt{learnr} is an open source R package that converts R Markdown documents
to programming tutorials using Shiny \citep{GarrettGrolemund2017}.  These
tutorials can consist of code exercises, multiple choice questions, videos,
and interactive Shiny components.  \texttt{learnr} enables instructors to
write their own programming tutorials, but there is not a collection of readily available
lessons.  Although the ``hint'' options in \texttt{learnr}
are minimal, the \texttt{checkr} package works with \texttt{learnr} to provide
tailored feedback \citep{DanielKaplan2018}.  \texttt{learnr} tutorials can be deployed either
locally on a user's computer (requires R installation) or through a Shiny Server (web-based).
This interactive tutorial does not allow for tracking of student progress.

@@

The simplest way to begin building the course is to use the template course files from
the \textit{Create Course} dialog. This will automatically create a GitHub repository
that is linked to DataCamp.

@@

Course creators can edit their courses through either GitHub or the \textit{Teach Editor}
accessed through the DataCamp website. The \textit{Teach Editor} allows content
creators to edit, preview, save changes, and synchronize to the GitHub repository.
Editing completed via GitHub does not allow for automatic content viewing. Any changes
that are made through the DataCamp \textit{Teach Editor} will update the GitHub
repository and vice versa.

@@

Figure \ref{fig:code1} is an example of ``normal exercise'' syntax and Figure
\ref{fig:preview} is the corresponding student interface. Students read the
lesson information and instructions, answer the question, and submit their code.

@@

\begin{figure}
\begin{Verbatim}[frame=single]
--- type:NormalExercise lang:r xp:100 skills:1 key:9a4fea2da0

@@

\end{Verbatim}
\caption{Syntax to create a ``normal exercise'' on DataCamp.  In DataCamp's
\texttt{Teach Editor}, the syntax is color-coded. \hl{Just include
screenshot here instead of verbatim so color coding is visible?}}
\label{fig:code1}
\end{figure}

\begin{figure}[h]
\includegraphics[width = 1.0\textwidth] {code1preview.png}
\caption{Preview of the rendered result of the ``normal exercise'' syntax displayed
in Figure \ref{fig:code1}.}
\label{fig:preview}
\end{figure}

@@

The SCT automatically assesses if the correct code was submitted \citep{Vankrunkelsven2016}. 
If a student submits an incorrect answer, they will receive immediate feedback and guidance towards
the correct answer.  The SCT is executed through DataCamp's \texttt{testwhat} R
package (\url{https://github.com/datacamp/testwhat/wiki}). The \textit{testwhat} package
contains functions that test different types of problems including multiple choice
exercises, what students typed, and output generated.  The SCT compares the ideal answer
(from the solution code) to the student's answer. Various arguments can be added to the
test functions to adjust the testing process and feedback.  Instructors can either use the
default feedback that DataCamp automatically generates or write their own custom messages.
For example, Table \ref{tab:SCT} contains the default feedback for a few possible incorrect submissions
of the \texttt{hist()} function from the example
exercise \textit{Visualizing quantitative data in R} shown in Figures \ref{fig:code1} and
\ref{fig:preview}.

\hl{I don't think I have fully utilized the power of SCT in my example because
you can link multiple SCTs together, right?  Could use a different example.}

\begin{table}
\begin{tabular}{lll}
\toprule
Student submission & R console & SCT feedback \\
\midrule
\texttt{histogram(CPS85\$wage)} & Error: could not find & Have you called hist()? \\
                                & function ``histogram'' & \\
[2ex]
\texttt{hist(wage)} & Error: object `wage' & Follow the format: hist(dataset\$variable) \\
                    &  not found           & with specified dataset and variable.\\
[2ex]
\texttt{hist(CP85\$wage)} & Error: object `CP85' & Follow the format: hist(dataset\$variable) \\
                               & not found       & with specified dataset and variable. \\
[2ex]
\texttt{hist(CPS85\$wage)} & & Great work! \\
\bottomrule
\end{tabular}
\caption{Feedback provided in the R console and in the DataCamp lesson through the
SCT for the exercise shown in Figures \ref{fig:code1} and \ref{fig:preview}.}
\label{tab:SCT}
\end{table}

@@

\begin{table}
\begin{tabular}{ll}
\toprule
Type & Description \\
\midrule
\texttt{VideoExercise}              & Displays a video exercise \\
\texttt{NormalExercise}	            & Instructions, exercise, code editor, and console \\
\texttt{MultipleChoiceExercise}     & Multiple choice question and console \\
\texttt{PureMultipleChoiceExercise} & Multiple choice question without a console \\
\texttt{TabExercise}	            & A series of connected sub-exercises \\
\texttt{BulletExercise}	            & A series of connected sub-exercises \\
\bottomrule
\end{tabular}
\caption{Exercise types currently supported in DataCamp}
\label{tab:exercises}
\end{table}

@@

The \texttt{MultipleChoiceExercise} contains a question and a list of various answers.
This exercise is accompanied by the R console and R output, and allows instructors to
pre-program objects or plots that the student needs to access to answer the question.
The \texttt{PureMultipleChoiceExercise} contains a question and a list of various
answers - this exercise type does not include the R console and output. 
%Similar to a
%\texttt{NormalExercise}, the lesson, instructions, hint, pre-exercise code and SCTs are
%included in the exercise syntax. 
Instructors write the questions in the lesson portion
of the exercise, present options in the \textit{Instructions} section, and can specify a
\textit{hint} as well. The \texttt{MultipleChoiceExercise} utilizes SCT for assessment,
whereas the \texttt{PureMultipleChoiceExercise} utilizes simpler syntax.

@@

Each DataCamp course is associated with a primary content creator, but multiple individuals
can contribute simultaneously to the course.  To do so, go to the course repository on GitHub
and under \textit{Settings} enter the Github user names of collaborators. One note of caution
here - to deploy the course to your class, the primary content creator needs to be listed
as member of the course in the DataCamp group.

@@

Integration of DataCamp assignments with Moodle was easy.  A link with the assigned
chapter was posted on the course site, and then students were taken directly to the
first page of the assigned chapter (no additional log-in required!). Grades were
available in the Moodle grade book immediately after students completed the assignment.
DataCamp computes scores as experience points (\texttt{xp}), but Moodle automatically
converted this a percentage completion as desired.  In addition, students were
able to re-visit the chapter as many times as they wanted.

@@

While there will always be some students (among non-statistics majors) that grumble about
having to learn R, overall student feedback was positive (see Table \ref{tab:evals}
in the Appendix for comparison of course evaluations with and without DataCamp).
Student comments included:
\begin{itemize}
\item \emph{DataCamp is really easy and fun to use. It is useful because it resembles the actual lab itself and we are still trying to figure out how to code ourselves but we are not penalize if we get the wrong answer.}
\item \emph{Data camp was helpful because it was actually like R.}
\item \emph{I think it's pretty good because it gives you an interactive way to look over what you'll need to do in class, but doesn't take super long. }
\end{itemize}
\hl{Clearly, I am cherry-picking here and there were criticisms of DataCamp as well. 
How best to present?}

@@

With integration via Moodle, one chapter was assigned at a time.  Although it
was possible to specify a recommended deadline for assignment completion, this
was not enforced via Moodle.  This meant that students could complete the
assignment after the deadline and still receive full credit.  Moreover,
because students are able to re-visit a chapter, the grade recorded in the Moodle
grade center was always the highest grade recorded for the student, regardless of
the completion iteration (first, second attempt?) or date of completion.  However,
this is more of a Moodle integration issue than a DataCamp issue.

@@

%--------------------------------------------------------------------------------
\subsection{Other possibilities}

Although we only discussed DataCamp classroom experiences for non-majors, this
platform is also well-suited and appropriate for undergraduate students in statistics 
and data science, as well as graduate students in statistics or even other disciplines. 
In addition, DataCamp can be a great learning tool for students
as they embark upon independent student research.

@@

While DataCamp is an incredible interactive tool to learn coding, there are some
pedagogical disadvantages. First, DataCamp can be fast-paced. Of course, the 
each user advances through the tutorial at their own speed; however, DataCamp
lessons are designed to keep moving forward with no clear pauses for reflection.
Another consequence of this is that DataCamp courses tend to assess correct
input of code, but not correct interpretation of results. Educators building
their own course content can incorporate interpretation questions based on 
achieved output; however, due to the modularity feature of DataCamp sessions
are not cumulative (every exercises begins with a clear workspace) so that 
an object created in a coding exercise would need to be re-defined for the student
in a multiple choice exercise. Lastly, due to the fast-paced nature, it is easy
for students to feel like they ``get it'' in the moment, but have little
recall as to what they did the next day. Educators can overcome this by 
encouraging or requiring note-taking, or by providing quick reference guides
of syntax learned for later use.

While DataCamp provides a platform to learn coding in R, it does not enable
users to fully utilize the reproducible and version control capabilities
of RStudio. Although it does contain courses such as 
\emph{Reporting with R Markdown} and \emph{Introduction to Git for Data Science},
it does not enable users to generate Markdown style reports from the web-based
interface. \hl{(Can you imagine if there was a Markdown button at the end of a lesson
to generate a report of what you learned?)} For this reason, educators may
want to use DataCamp as a learning aid, but not as a primary mode for 
student activities.

DataCamp is a private, for-profit company that has gracefully allowed
the academic community to utilize their resources for free.  However, 
this means that both the website and way in which lesson content
is created is constantly changing.  For example, between September of
2017 and April of 2018, DataCamp:
\begin{itemize}
\item Changed exercise headers from single text line to a \texttt{yaml}
\item Changed the naming of the exercise \texttt{PlainMultipleChoice} 
to \texttt{PureMultipleChoice}
\item Added new exercise options (the connected sub-exercises)
\item Changed the \emph{teach} documentation
\item Changed the web-site interface for \emph{groups}
\end{itemize}
While rapid and continuous development of DataCamp products is beneficial 
to both educators and learners, it can be jarring for a new educator
who just got accustomed to the platform.  Moreover, community
content creators are not notified when these changes are deployed.  
Thankfully, for now, lessons created under older formats (i.e.,
with \texttt{PlainMultipleChoice} or headers in a single text
line instead of a \texttt{yaml}) still render correctly and no updates are required to
lesson syntax. This also means that any specific weblinks listed
here or exercise syntax may already have changed by the time you read this.

@@

DataCamp fills a much needed void for interactive coding exercises that allow
for tracking of student progress.  Due to the breadth of courses already offered as
well as the ability to create your own course content, DataCamp is a learning
platform that can be used by any instructor that incorporates coding into their
course content.  \hl{A bit more here?}

%================================================================================
\section{Appendix}
\begin{table}
 \resizebox{\textwidth}{!}{ 
\begin{tabular}{lcc}
\toprule
	&	Winter 2017	  &	Winter 2018	\\
    &   $(n = 63)$    &  $(n = 59)$  \\
\midrule
\multicolumn{3}{l}{\textbf{How do you rate the following items in terms of building your knowledge of introductory statistics?}} 					\\
\multicolumn{3}{l}{\textbf{1 = Not valuable/ did not contribute; 4 = very valuable/highly contributed}}					\\
Lab videos	            &	3.24	&   N/A		\\
DataCamp assignments	&	N/A	    &	2.47	\\
Lab assignments	        &	2.86	&	2.81	\\
\midrule
\multicolumn{3}{l}{\textbf{How well did the course address the following learning objectives?}}					\\
\multicolumn{3}{l}{\textbf{1 = poor; 5 = great}}					\\
Use R as a tool to perform statistical analysis	&	3.84	&	3.90	\\
\midrule
\multicolumn{3}{l}{\textbf{How were the pre-lab and lab assignments?}}					\\
\multicolumn{3}{l}{\textbf{1 = strongly disagree; 5 = strongly agree}}					\\
I had difficulty understanding the connections between lecture and lab.             	&	2.60    &	2.88	\\
The pre-lab assignments sufficiently prepared me for lab	                            &	4.00	&	3.80	\\
The lab assignments helped me see how statistical analysis is done in the real world.	&	3.63	&	3.51	\\
DataCamp was easy to use	                                                            &	N/A	    &	4.15	\\
I enjoyed learning how to code on DataCamp	                                            &	N/A	    &	3.32	\\
\midrule
\multicolumn{3}{l}{\textbf{How did you like using R in this class?}}					\\
\multicolumn{3}{l}{\textbf{1 = strongly disagree; 5 = strongly agree}}					\\
R was challenging to use the entire term	                    &	3.05	&	2.76	\\
R became easier to use as the term progressed	                &	4.08	&	4.00	\\
Using R made it hard to learn how to do statistical analysis	&	2.46	&	2.32	\\
Using R helped me understand how to do statistical analysis	    &	3.17	&	3.39	\\
I enjoyed using R	                                            &	2.87	&	2.92	\\
I am glad I learned R	                                        &	3.46	&	3.61	\\
\bottomrule
\end{tabular}}
\caption{Course evaluations from two sections of STAT 217 in Winter 2017 (without DataCamp, response rate = 90\%) 
and Winter 2018 (with DataCamp, response rate = 82\%) taught
by the same instructor with otherwise virtually no changes.}
\label{tab:evals}
\end{table}

\hl{Is this table even worthwhile or of interest to include?  If so, I can work on better formatting.  Include
SDs?  Present graphically instead?}

%================================================================================
\bibliographystyle{plainnat}
\bibliography{library}
